---
title: "Milestone 4"
author: vub23
format: pdf
---

```{r setup, include = F}
# Add packages you want here bru
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.height = 10)
options(scipen = 7)
library(tidyverse)
library(stargazer)
library(corrplot)
library(patchwork)
library(tigris)
library(sf)
library(factoextra)
library(kableExtra)
```

This project will implement prof Sage's recommendations for the new scoring paradigm (subjective ranking informed by stats, score curve interpolation, time and applying the curve on the census tracts). Any nontrivial step will get documented for each variable. I aim to get 5 done for this milestone.

Preliminary things:

-   Our census tracts will be compared against national statistics (i.e. 0 would be close to the worst tracts in the country while 100 would be close to the best)

-   The 5 thresholds for "ideal", "good", "average", "bad" and "terrible" are all subjectively set based on available information. This means that aside from the lack of data for some variables (national data bodies will most likely not have data to the same level of detail), we might handle some variables in questionable/contentious ways.

-   Everything is subject to change. Jason's opinion will determine how we do the other variables/categories.

-   What we have done for milestones 2 and 3 are in the dumpster. We will start from scratch.

## Variable 1: "Rent"

Jumping headlong into stuff is my favorite thing to do, how could you tell?

Absolute rent levels mean nothing if we are talking about a national scale - the sole addition of Manhattan's tracts will throw our entire proposed scale off balance. If this is the case, we should probably think of something a little more income-based. Using 2 variables: median housing costs `MHC` and median household income `INC` (from a new hh census tracts file), we can find the housing cost to income ratio.

Some searches reveal that US' best cities tend to have a rent-to-income ratio of \~13% while the worst ones (New York, Miami, Seattle) would be way north of 50%, even 60%. Knowing that MHC is inclusive of mortgages, but also other fees and taxes, I would push this estimate a little to the right: our "ideal" would be any tract with the ratio less than 15%, our "terrible" would be any tract with a ratio of 50% or higher. 30% is proposed as the highest proportion of monthly income the average person should spend on rent, so we would leave it as our "average". That leaves us with 23% as the "good" benchmark and 40% as the "bad" threshold. Now to incorporate that into the dataset:

```{r}
# Read csv file
hh <- read.csv("hh tracts 2.csv")
hh <- hh %>% select(-Layer, -Name)

# Copying Mina's code for the string pruning process
names(hh) <- sub("_.*$", "", names(hh))

# Creating other dataframes that fit the requirements
hhCorr <- hh %>% select(-HCP) %>% na.omit()
hhSumm <- hh %>% select(-GEOID)

# Creating another column calculating the ratio:
hh <- hh %>% mutate(icPct = MHC/(INC/12)*100, icScore = approx(x=c(15,23,30,40,50), y=c(100,75,50,25,0), xout=icPct, rule=2)$y)

# A lot happened in that 1 chunk, so I'll explain here. xout is the column you feed the raw data into. approx() is the function that does the linear interpolation thing. x=() is the argument where you fit the ideal/terrible thresholds. y=() is the argument where you assign the values of each threshold. rule=2 is set so values beyond the bounds are either 100 or 0.
```

Let's get a table showing the distribution of the new score:

```{r, results='asis'}
score1 <- hh %>% select(icScore)
stargazer(score1, type = 'text', summary.stat = c("n","mean","sd", "min", "median", "max"), notes.append=F, header=F)
```

A median value of almost 96 –\> Very right skewed. However, this tracks with our expectations of Wisconsin/the Midwest in general. We do have very affordable housing in comparison with other areas in the US.

## Variable 2: Eviction Rates

I'd also argue that this is a very important variable in terms of humane housing, though there are challenges working with eviction rates we must account for:

-   Eviction can mean many things: it can be the filing of the legal process that removes tenants from a property, it can also be the actual removal of such residents from their homes.

-   Some jurisdictions allow landlords to file multiple eviction notices, strongly skewing extreme values (some Maryland counties report a 147% eviction rate, for example).

-   However, the bulk of values (both external and from the dataset) seem to be on the low end, so the thresholds will reflect that.

Currently, the thresholds for ideal/good/average/bad/terrible are: 0%/1%/3%/7%/13%. Below is a table for how the score for the variable looks:

```{r, results='asis'}
hh <- hh %>% mutate(evScore = approx(x=c(0,1,3,7,13), y=c(100,75,50,25,0), xout=EVR, rule=2)$y)
score2 <- hh %>% select(evScore)
stargazer(score2, type = 'text', summary.stat = c("n","mean","sd", "min", "median", "max"), notes.append=F, header=F)
```

## Variable 3: Home Ownership Rates

As noted by Jason, home ownership rates is a very important metric for humane housing, partly because it's a nice proxy to more humane conditions. There are still some considerations regarding this variable though:

-   Higher `HUO` rates might indicate higher neighborhood stability, but also lack of rental housing.

-   Tracts and areas with transient populations (like dorms and institutional quarters) might have disproportionately low levels of `HUO`, and therefore should be addressed appropriately.

```{r}
hh <- hh %>% mutate(hoScore = case_when(
      HUO < 20 ~ NA_real_,
      TRUE ~ approx(
        x    = c(20, 40, 60, 80, 90),
        y    = c(0, 20, 50, 80, 100),
        xout = HUO,
        rule = 2
      )$y))
score3 <- hh %>% select(hoScore)
stargazer(score3, type = 'text', summary.stat = c("n","mean","sd", "min", "median", "max"), notes.append=F, header=F)
```

## Variable 4: Mean Travel Time

This variable seems to be the most straightforward to create an index from. Some research has confirmed that any commute less than 16 minutes is ideal, and 60+ commutes to be harmful, so we know where to place our thresholds. Here is another table that shows the distribution of our score:

```{r}
hh <- hh %>% mutate(trScore = approx(
        x    = c(16, 20, 25, 35, 60),
        y    = c(100,  80, 50, 20,0),
        xout = TRV,
        rule = 2
      )$y)
score4 <- hh %>% select(trScore)
stargazer(score4, type = 'text', summary.stat = c("n","mean","sd", "min", "median", "max"), notes.append=F, header=F)
```

## Variable 5: Diversity

The last variable explicitly mentioned by Jason is that of racial/ethnic diversity. Though arguments can be made against our current prior that more diversity is conducive to more humane housing, we are still pursuing such a paradigm. Knowing that the highest possible value is 0.875, we should turn to other sources of data and examine the ranges of possible values.

The biggest, most diverse cities in the country like New York and LA tend to have a `REX` of 0.75 and higher. Conversely, counties like Laredo, Texas, with a predominantly hispanic population would only come in at 0.1 or lower on the scale. Given that this is the case, my current suggestion for the thresholds are 0.15, 0.25, 0.5, 0.65 and 0.75 for conditions ranging from terrible to ideal. Below is a table showing us the distribution of `REX` scores across different tracts of our dataset.

```{r, results='asis'}
hh <- hh %>%
  mutate(diScore = approx(
    x    = c(0.15, 0.25, 0.50, 0.65, 0.75),
    y    = c(0,    20,   50,   80,  100),
    xout = REX,
    rule = 2
  )$y)
score5 <- hh %>% select(diScore)
stargazer(score5, type = 'text', summary.stat = c("n","mean","sd", "min", "median", "max"), notes.append=F, header=F)
```

## Putting it all together

Here is a map containing all the census tracts with the new scoring paradigm:

```{r}
# Arbitrary limits decided by Mina (so, not arbitrary)
zoom_xlim <- c(-89, -88)
zoom_ylim <- c(43.8, 44.7)

ind <- hh %>%  select(GEOID, ends_with("Score")) %>%  
  rowwise() %>%  mutate(newIndex = mean(c_across(ends_with("Score")), na.rm = TRUE)) %>% 
  ungroup() %>%  arrange(desc(newIndex))

# Getting spatial data from tigris, zcta was found from reading the documentation
zips <- unique(hh$GEOID)
spatial <- tracts(state="wi", class="sf", year=2023, progress_bar=F)
foxZips <- spatial %>% mutate(GEOID=as.character(GEOID)) %>% filter(GEOID %in% zips)
# Getting other zips from tigris so that we know where we are
background <- spatial %>%
  mutate(GEOID=as.character(GEOID)) %>%
  filter(!GEOID %in% zips)

# joining the data tables together
ind <- ind %>% mutate(GEOID=as.character(GEOID)) %>%  left_join(foxZips, by=c("GEOID"="GEOID"))

ind <- ind %>% st_as_sf()


# Drawing up the map
ggplot() +
  geom_sf(data = ind, aes(fill = newIndex), color="black") +
  geom_sf(data = background, fill = "grey95", color = "grey80", linewidth = 0.1) +
  scale_fill_viridis_c() + 
  labs(title = "Map of Zipcodes by Composite Housing Index") +
  theme_minimal() + coord_sf(xlim = zoom_xlim, ylim = zoom_ylim, expand = FALSE)
```

Yay!
