---
title: "idk"
author: cheese
format: pdf
---

```{r setup, include=F}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, fig.height = 10)
options(scipen = 7)
library(tidyverse)
library(stargazer)
library(corrplot)
library(patchwork)
library(tigris)
library(sf)
library(factoextra)
```

We look into lifelong learning for the census tract subdivision.

```{r}
#file orwhatever
# getting the data into readable format
colnames <- names(read.csv("Lifelong-Learning-Cencus-Tracts.csv",nrows=0))

# data wrangling blah blah blah
balls <- read.csv("Lifelong-Learning-Cencus-Tracts.csv", skip=2, header=FALSE)
names(balls) <- colnames
balls <- balls |> select(-Layer) # removing redundant columns

# summary stats dataset
ballsSumm <- balls |> select(-GEOID, -Name)
```

## Legend

Similar to the zip code level, the census tract variable names do not provide much insight into what they are representing. Thus, I will describe them here. Please note that all of this data was collected from 2019-2023.

- `EDA`: 9th grade education rate (% of residents)
- `EDB`: High school graduation rate (% of residents)
- `EDC`: Any higher education rate (% of residents)
- `EDE`: College graduation rate (% of residents)
- `EDG`: Preschool enrollment rate (% of toddlers ages 3-4)

Good news! There are no `NA` entries in this dataset at all, so we do not need to worry about that!!

## Exploratory Plots and Variables

### Correlation Plots


### Histograms


### Summary Stats


## Normalization and Z-score calculations

None of the variables need to be flipped, as higher education/graduation rates implies a more educated population, which is generally viewed as a good thing.

```{r}
# normalization and zscore calc here
```


## Indexing and Visualization



