---
title: "DeyWork"
format:
  pdf:
    documentclass: article
    include-in-header:
      text: |
        \usepackage[margin=0.75in]{geometry}

---

As a Disclaimer, much of the code for this first draft is from Binh and Mina's work. Thanks Binh and Mina!
```{r}
library(tidyverse)
library(dplyr)
library(knitr)
library(kableExtra)
library(patchwork)
library(corrplot)
library(tinytex)
library(purrr)
library(ggplot2)
library(rlang)
library(tidyr)
```

```{r, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.width = 7)
```


```{r}
# read in the csv
transportationZIP <- read_csv("TransportationZIPCodes.csv") 

```


```{r}
# first look at the dataset
glimpse(transportationZIP)
```
## Short Description of the Variables

GEOID: ZIP code

Population: total number of residents

ACT_2019-2023: % of workers who walk or bike to work

CAR_2019-2023: % of workers who drive alone

PUB_2019-2023: % of workers using public transit

NVC_2019-2023: % of households without a vehicle

TRV_2019-2023: mean travel time to work (minutes)

EKW_2024: Walkability Index
Longitude, Latitude: spatial coordinates (not used in index)



```{r}
# remove first 4 rows 
transportation <- transportationZIP [-c(1:4), ]
```

```{r}
# Name and GEOID columns seem to be Identical 

## Drop the Name column
transportation <- transportation %>% select(-Name)
```

```{r}

# 2. Convert some columns to numeric
num_cols <- c(
  "GEOID","Population", "Longitude", "Latitude", 
  "ACT_2019-2023", "CAR_2019-2023", "EKW_2024",
  "NVC_2019-2023", "PUB_2019-2023", "TRV_2019-2023"
)

transportation <- transportation %>%
  mutate(across(all_of(num_cols), ~ as.numeric(.)))

```

```{r}
# I removed 1 row that contained NA values
# There was only one row which had NA values for the 54985 ZIP code. This is in the Winnebago Area  
transportation <- transportation %>% drop_na()
```

## Summary Statistics
```{r}

summary_stats <- transportation %>%
  # Convert all character columns to numeric safely
  mutate(across(where(is.character), ~ suppressWarnings(as.numeric(.x)))) %>%
  select(-Layer, -GEOID, -Longitude, -Latitude) %>%
  summarise(across(
    everything(),
    list(
      count  = ~ sum(!is.na(.x)),
      mean   = ~ mean(.x, na.rm = TRUE),
      median = ~ median(.x, na.rm = TRUE),
      sd     = ~ sd(.x, na.rm = TRUE),
      min    = ~ min(.x, na.rm = TRUE),
      max    = ~ max(.x, na.rm = TRUE),
      p25    = ~ quantile(.x, 0.25, na.rm = TRUE),
      p75    = ~ quantile(.x, 0.75, na.rm = TRUE)
    ),
    .names = "{.col}_{.fn}"
  )) %>%
  
  # Split only at the LAST underscore using regex
  pivot_longer(
    cols = everything(),
    names_to = c("Variable", "Statistic"),
    names_pattern = "^(.*)_(\\w+)$",
    values_to = "Value"
  ) %>%
  pivot_wider(names_from = Statistic, values_from = Value)

# Display clean summary table
summary_stats %>%
  kable(caption = "Summary statistics for transportation indicators", digits = 2, booktabs = TRUE) %>%
  kable_styling(full_width = FALSE, position = "center")


```

## Histograms

```{r histograms, fig.cap="Histograms of transportation variables"}

# Variables to plot (in quotes exactly as they appear in the data)
vars <- c("ACT_2019-2023", "CAR_2019-2023", "PUB_2019-2023",
          "NVC_2019-2023", "TRV_2019-2023", "EKW_2024")

# Plot histograms using patchwork layout
wrap_plots(
  map(vars, function(v) {
    ggplot(transportation, aes(x = .data[[v]])) +
      geom_histogram(bins = 15, fill = "#3182bd", color = "black") +
      labs(title = v, x = NULL, y = "Count") +
      theme_minimal(base_size = 10)
  }),
  ncol = 3
)

```

ACT_2019-2023 (active transportation) is right-skewed, indicating that most ZIP codes have very low rates of walking/biking to work.

CAR_2019-2023 shows a tight, slightly left-skewed distribution, meaning most areas have high levels of car usage.

PUB_2019-2023 is highly right-skewed: public transit use is nearly zero in most areas.

NVC_2019-2023 (no vehicle) and TRV_2019-2023 (commute time) are moderately skewed.

EKW_2024 (walkability) is fairly normal, with most ZIP codes clustered around the median but some with higher walkability.

Insight: Some variables (like public transit) may be more informative in only a few ZIP codes, suggesting that weighting or transformation may be necessary to prevent them from distorting the index.

## Correlation Matrix
```{r corr-matrix, fig.cap="Correlation Matrix of Indicators"}
corr_data <- transportation %>%
  select(
    `ACT_2019-2023`, `CAR_2019-2023`, `PUB_2019-2023`,
    `NVC_2019-2023`, `TRV_2019-2023`, EKW_2024
  ) %>%
  mutate(across(everything(), as.numeric))

corrplot(
  cor(corr_data, use = "pairwise.complete.obs"),
  method = "color", type = "upper",
  order = "AOE", addCoef.col = "black")
```

The correlation matrix shows a strong positive relationship between non-vehicle commuting (NVC) and active transportation (ACT) (r = 0.65), suggesting that areas with more biking/walking also tend to have higher non-car commuting.

Walkability (EKW_2024) is moderately correlated with NVC (r = 0.54) and public transportation use (PUB) (r = 0.50), supporting the idea that walkable areas facilitate alternative commuting methods.

Travel time (TRV) is negatively correlated with walkability (r = -0.61) and public transit use (r = -0.42), suggesting that more walkable, transit-friendly areas may reduce commute durations.

Car commuting (CAR) shows expected negative correlations with ACT (r = -0.39) and NVC (r = -0.31), reinforcing that higher car reliance is associated with lower use of alternative modes.

These results indicate redundancy between ACT, NVC, and walkability, which should be considered when selecting and weighting variables for index construction.

\newpage

## Z Scores & Index Construction

```{r}
# Flip direction so that larger = "better"
trans_good <- transportation %>% 
  mutate(
    across(
      c(`CAR_2019-2023`, `NVC_2019-2023`, `TRV_2019-2023`),
      ~ -.x                     # negate “bad-when-high” vars
    )
  )


```

```{r}
#Standardise to z-scores
score_vars <- c("ACT_2019-2023","CAR_2019-2023","PUB_2019-2023",
                "NVC_2019-2023","TRV_2019-2023","EKW_2024")

trans_z <- trans_good %>% 
  mutate(across(all_of(score_vars),
                ~ (as.numeric(.x) - mean(as.numeric(.x), na.rm = TRUE)) /
                  sd(as.numeric(.x), na.rm = TRUE),
                .names = "{.col}"))


```

```{r}
# Equal-weight composite index
index_tbl <- trans_z %>% 
  mutate(Index = rowMeans(select(., all_of(score_vars)), na.rm = TRUE)) %>% 
  arrange(desc(Index))


```


```{r}
# Table of z-scores + index
index_tbl %>% 
  select(GEOID, all_of(score_vars), Index) %>% 
  kable(caption = "Z-scores and equal-weight transportation index",
        digits = 2, booktabs = TRUE) %>% 
  kable_styling(full_width = FALSE, position = "center", font_size = 9)


```

```{r fig.cap="Composite Transportation Index (equal weights)"}
# Bar plot of index scores 
ggplot(index_tbl, aes(x = reorder(GEOID, Index), y = Index)) +
  geom_col(fill = "#31a354") +
  coord_flip() +
  labs(x = "ZIP code", y = "Index (z-score units)") 
```
As Binh said this plot does not give very meaningful insights but we can still take a look at it.
54901, 54911, and 54927 have the highest index scores, suggesting strong performance across variables like active transport, walkability, and low car use.
54152, 54980, and 54170 rank lowest, indicating poor transportation diversity or infrastructure.


# Map
```{r qtransport-map, fig.width=10, fig.height=10, fig.cap="ZIP codes ranked by composite transportation index"}

# Adapted from Binh's humane-housing map code. Thank you Binh!

library(tigris)   # spatial ZIP shapes
library(sf)       # simple-features ops
library(ggplot2)  # plotting
options(tigris_use_cache = TRUE, tigris_class = "sf")

# Pull WI ZIP polygons
zips_needed <- unique(index_tbl$GEOID)
wi_zctas     <- zctas(state = "wi", class = "sf", year = 2010, progress_bar = FALSE)

map_zips  <- wi_zctas %>% 
  mutate(ZIP = as.character(ZCTA5CE10)) %>% 
  filter(ZIP %in% zips_needed)

background <- wi_zctas %>% 
  mutate(ZIP = as.character(ZCTA5CE10)) %>% 
  filter(!ZIP %in% zips_needed)

# Join index scores to spatial layer
transport_sf <- index_tbl %>% 
  mutate(GEOID = as.character(GEOID)) %>% 
  left_join(map_zips, by = c("GEOID" = "ZIP")) %>% 
  st_as_sf()

# Plot
zoom_xlim <- c(-89.5, -87.5)   # same crop Binh used
zoom_ylim <- c(43.5, 45.0)


# Added text with the zip codes, some dont align perfectly but i thought this is helpfull
ggplot() +
  geom_sf(data = background, fill = "grey95", color = "grey80", linewidth = 0.1) +
  geom_sf(data = transport_sf, aes(fill = Index), color = "black") +
  geom_sf_text(data = transport_sf, aes(label = GEOID), size = 2.5, color = "white")+
  scale_fill_viridis_c(name = "Index") +
  coord_sf(xlim = zoom_xlim, ylim = zoom_ylim, expand = FALSE) + 
  labs(title = "Composite Transportation Index by ZIP Code") +
  theme_minimal()
```

High-scoring ZIPs are concentrated in urban centers (e.g., Oshkosh, Appleton).
Lower scores dominate in rural or exurban areas, consistent with lower population density, fewer sidewalks, and car dependency.

